---
title: "Assigment on dimension reducion"
author: "Rafał Rysiejko, 423827"
date: "12/23/2019"
output:
  rmarkdown::pdf_document:
    fig_caption: yes        
    includes:  
      in_header: my_header.tex
      
---

```{r setup, include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

The recorded music industry was worth $19.1 billion in 2018, which was almost a double-digit gain (9.7%) from the year prior[^1]. In this entertainment market, an increasingly large number of song products are introduced each year. However, only a small group achieves mainstream success, and among these, the distribution of market success is becoming increasingly skewed. By analyzing the top songs from the popular music streaming service provider, we can investigate if some underlying patterns make a song truly hit.  

This assignment will try to determine whether there are some distinctive features of song that make it successful and if so, are there any patterns among those features. For this, a set of statistical methods will be used, including dimensionality reduction techniques.
Dataset[^2] used for this assignment contains information about the top 50 most listened songs in the world on music streaming platform Spotify in 2019. Each song has a set of 13 variables:  

   
\setlength{\leftskip}{6cm}

Table 1: Variable description

\setlength{\leftskip}{0pt} 


```{r table2, echo=FALSE, message=FALSE, warnings=FALSE, results='asis',fig.align = "center"}
tabl <- "
| Variable    | Description         |
|-----------|:------------------------| 
| Track.Name          |Song Title |
| Artist.Name         | Artist performing the song |
| Genre               | The genre of the track   |
| Beats.Per.Minute    | Variable describing the tempo of the song.  |
| Energy              | The energy of a song - the higher the value, the more energtic song.    |
| Danceability        | The higher the value, the easier it is to dance to this song.    |
| Loudness..dB.       | The higher the value, the louder the song.   |
| Liveness            | The higher the value, the more likely the song is a live recording.    |
| Valence             | The higher the value, the more positive mood for the song.  |
| Length              | The duration of the song. |
| Acousticness        | The higher the value the more acoustic the song is. |
| Speechiness         | The higher the value the more spoken word the song contains.  |
| Popularity          | The higher the value the more popular the song is. |
"
cat(tabl) 
```


## Exploratory data analysis

Installing and running the libraries:
```{r echo=T, message=FALSE, warning=FALSE, results='hide'}
requiredPackages = c("tidyverse","factoextra","stats","clustertend","flexclust","ggforce"
,"fpc","cluster","ClusterR","knitr","kableExtra","DataExplorer","caret",
"reshape2","corrplot","labdsv","smacof","clusterSim","pastecs","psych","pca3d","pls") 
for(i in requiredPackages){if(!require(i,character.only = TRUE)) install.packages(i)} 
for(i in requiredPackages){library(i,character.only = TRUE) } 
```

Loading the data:
```{r echo=T}
data_full <- read.csv("top50.csv",stringsAsFactors = F)
data_full <- data_full[,2:14]
data_init <- data_full
```

```{r echo=F}
x1 <- t(round(stat.desc(data_init[,c(4:13)]),2))
x1 <- x1[,c(4:6,8:13)]

kable(x1,caption = "Summary statistics of the dataset")%>% 
kable_styling(latex_options="hold_position",bootstrap_options = c("striped", "hover"))
```

Songs in the data set are grouped into `r length(unique(data_init$Genre))` genres.
 
```{r echo=F,message=FALSE,out.width = '70%',fig.align = "center",fig.cap="\\label{fig:figs}Genre distribution."}
data_init %>% count(Genre) %>%
  ggplot(aes(reorder(Genre, n), n)) + geom_col(fill="cyan3") + coord_flip() + ylab("Count") +xlab("") + theme_minimal()
```
 
As there a lot of subgenres with a low number of instances, a new variable `aggregate.genres` is created.
  
  
```{r echo=T,message=FALSE}
data_init$aggregate.genre <- NA
data_init$aggregate.genre <- ifelse(data_init$Genre  %in% c("canadian pop","pop","dance pop",
                        "electropop","panamanian pop","pop house","australian pop"),"pop",
ifelse(data_init$Genre  %in% c("dfw rap","country rap","canadian hip hop","atl hip hop"),"hip hop",
ifelse(data_init$Genre  %in% c("r&b en espanol","latin"),"latin",
ifelse(data_init$Genre  %in% c("reggaeton flow","reggeaeton"),"reggaeton",
ifelse(data_init$Genre  %in% c("edm","trap music","brostep"),"electronic",
ifelse(data_init$Genre  %in% c("escape room","boy band","big room"),"other",data_init$Genre))))))

data_init$aggregate.genre <- as.factor(data_init$aggregate.genre)
``` 
 
  
  
```{r echo=F,message=FALSE,out.width = '70%',fig.align = "center",fig.cap="\\label{fig:figs} Aggregated genre distribution."}
data_init %>% count(aggregate.genre) %>%
  ggplot(aes(reorder(aggregate.genre, n), n)) + geom_col(fill="cyan3") + coord_flip() + ylab("Count") +xlab("") + theme_minimal()
```
 

  
```{r echo=F,message=FALSE,out.width = '80%',fig.align = "center",fig.cap="\\label{fig:figs} Histograms of numerical variables."}
gather(data_init[,c(4:13)]) %>% 
  ggplot(., aes(value)) + 
  geom_histogram(aes(y =..density..), 
                 col="black", 
                 fill="lightblue", 
                 alpha=.2) + 
  geom_density(col="cyan3")+
  facet_wrap(~key, scales = 'free') + theme_minimal()
```

Variables `Liveness` and `Speechiness.` are highly positively skewed, suggesting the existence of outliers. A log transformation is applied to them as the PCA algorithm is sensitive to them.

```{r echo=T,message=FALSE}
transformed_var <- c("Liveness", "Speechiness.")
data_init <- data_init %>% mutate_at(vars(transformed_var), log)
``` 

Distribution of variables `Liveness` and `Speechiness.` after log transformation
  

To further investigate the relationship between variables, a correlation analysis was carried using a Pearson correlation coefficient.  


```{r echo=FALSE, message=FALSE,out.width = '50%',fig.align = "center",fig.cap="\\label{fig:figs} Correlation matrix."}

data_corr<- select_if(data_init, is.numeric)
corrplot::corrplot(cor(data_corr), type="lower", method="number",order ="alphabet")

```
  
  
From Figure 4, we can see a high positive correlation between variables `Speechiness` and `Beats.Per.Minute`, typical for rap songs. Another relationship is visible between variables `Energy` and `Loudness.dB.` suggesting that highly energetic songs tend to be louder. Also, a more energetic song, on average less acoustic.



## Principal Component Analysis.
  
The data set is first split into train and test subsets for following PCA regression purposes with an 80% to 20% ratio.  
  
  
```{r echo=T, message=FALSE}
smp_size <- floor(0.80 * nrow(data_init))

set.seed(123)
train_ind <- sample(seq_len(nrow(data_init)), size = smp_size)

data <- data_init[train_ind, ]
data_test <- data_init[-train_ind, ]
``` 

  
As different variables in the data set have different units of measurement, a normalization is required.  


```{r echo=T,message=FALSE}
data_pca <- data_init %>%  dplyr::select(-Popularity) %>% select_if(is.numeric)
data_pca_stand<-data.Normalization(data_pca, type="n1", normalization="column") 
``` 

\pagebreak 
Performing PCA
  
```{r echo=T,message=FALSE}
pca1<-prcomp(data_pca_stand, center=TRUE,scale. = TRUE)
x2 <- round(summary(pca1)$importance,2) 
``` 


### Summary of PCA  
  

```{r echo=F}
kable(x2,caption = "Summary statistics of the PCA") %>% 
kable_styling(latex_options = "hold_position",bootstrap_options = c("striped", "hover"),)
```
  
Based on the summary in table 3, we see that 56 percent of the variance is contributed by the first three principal components. We can also see that there is no significant decrease in explained variance with each variable, suggesting that each of them contributes significantly to the explained variance.  

  
```{r echo=FALSE, message=FALSE,out.width = '50%',fig.align = "center",fig.cap="\\label{fig:figs}Scree Plot."}
fviz_eig(pca1)
```

      
```{r echo=F, message=FALSE}
# Helper function 
#::::::::::::::::::::::::::::::::::::::::
var_coord_func <- function(loadings, comp.sdev){
  loadings*comp.sdev
}
# Compute Coordinates
#::::::::::::::::::::::::::::::::::::::::
loadings <- pca1$rotation
sdev <- pca1$sdev
var.coord <- t(apply(loadings, 1, var_coord_func, sdev)) 

# Compute Cos2
#::::::::::::::::::::::::::::::::::::::::
var.cos2 <- var.coord^2

# Compute contributions
#::::::::::::::::::::::::::::::::::::::::
comp.cos2 <- apply(var.cos2, 2, sum)
contrib <- function(var.cos2, comp.cos2){var.cos2*100/comp.cos2}
var.contrib <- t(apply(var.cos2,1, contrib, comp.cos2))
```

```{r echo=F}
kable(round(var.contrib[, 1:9],2),caption = "Contribution of a variable to a given principal component") %>% 
kable_styling(latex_options = "hold_position",bootstrap_options = c("striped", "hover"),)
```

  
```{r echo=FALSE, message=FALSE,out.width = '50%',fig.align = "center",fig.cap="\\label{fig:figs}Variable contribution to Principal Component "}
fviz_pca_var(pca1,
             col.var = "contrib", # Color by contributions to the PC
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE     # Avoid text overlapping
) + theme_minimal()
```

Figure 6 represents the contribution of variables to the first two Principal Components. Positively correlated variables point to the same side of the plot. Negatively correlated variables point to opposite sides of the graph.  
  
  
```{r echo=F, message=FALSE}
# Helper function 
#::::::::::::::::::::::::::::::::::::::::
var_coord_func <- function(loadings, comp.sdev){
  loadings*comp.sdev
}
# Compute Coordinates
#::::::::::::::::::::::::::::::::::::::::
loadings <- pca1$rotation
sdev <- pca1$sdev
var.coord <- t(apply(loadings, 1, var_coord_func, sdev)) 

# Compute Cos2
#::::::::::::::::::::::::::::::::::::::::
var.cos2 <- var.coord^2

# Compute contributions
#::::::::::::::::::::::::::::::::::::::::
comp.cos2 <- apply(var.cos2, 2, sum)
contrib <- function(var.cos2, comp.cos2){var.cos2*100/comp.cos2}
var.contrib <- t(apply(var.cos2,1, contrib, comp.cos2))
```

    
```{r echo=FALSE, message=FALSE,out.width = '70%',fig.align = "center",fig.cap="\\label{fig:figs}Graph of individual songs by aggregated genre."}
fviz_pca_ind(pca1,label="none",
             habillage=data_init$aggregate.genre,
             addEllipses=TRUE) + theme_minimal()
```

Figure 7 represents the position of individual songs on the space of principal components, where songs with a similar profile are grouped together. We can see that the songs from the same genre tend to be placed near each other, except for `pop`, suggesting that this genre has high variability in analyzed features.   
  

\pagebreak
### Rotated PCA using varimax rotation.  

```{r echo=T,message=F,warning=F}
n = 3
pca_rot <- principal(data_pca_stand, nfactors=n, rotate="varimax")
x3 <- pca_rot
```

```{r echo=F}
kable(round(x3$loadings[,1:n],2),caption = "Output 3") %>% 
kable_styling(latex_options = "hold_position",bootstrap_options = c("striped", "hover"),)
```
  
When analyzing the rotated components, we can distinguish the following composite features:

* First rotated component RC1 – Loudness and Energy -> dance club readiness
* Second rotated component RC2 – Beats per minute and Speechiness -> lyrical rhythm
* Third rotated component RC3 – danceability and liveness -> calm and concert ready 


  
```{r echo=F}
kable(round(cbind(pca_rot$complexity,pca_rot$uniqueness),2),caption = "Quality measures of th PCA",col.names = c("Complexity","Uniqueness")) %>% 
  kable_styling(latex_options = "hold_position",bootstrap_options = c("striped", "hover"),)
```
  
Table 8 represents two quality measures *Complexity*, which indicates how many variables constitute a single factor. High complexity is an undesirable feature because it implies a more difficult interpretation of factors. *Uniqueness*, on the other hand, explains the proportion of variance that is not shared with other variables, meaning that variable with lower value carries less additional information in relation to other variables in the model.  
  
  
### Principal components regression 

```{r echo=T}
data_pca_regression <- data.Normalization(data_init %>% select_if(is.numeric), 
type="n1", normalization="column")

reg.pcr.train<- pcr(Popularity~., data=data_pca_regression, scale=TRUE, 
                    center = TRUE,validation="CV",ncomp = 3)

data_test_regression <- data.Normalization(data_test 
      %>% select_if(is.numeric), type="n1", normalization="column")

reg.pcr.pred<-predict(reg.pcr.train, data_test_regression, ncomp = 3)
```
  

```{r echo=T}
summary(reg.pcr.train)  
```  
  
The summary  provides the percentage of variance explained in the predictors and the outcome (Popularity) using different numbers of components
For example, 56.27%% of the variation (or information) contained in the predictors is captured by three principal components (ncomp = 3). Additionally, setting ncomp = 3 captures only 9.67% of the information in the outcome variable (Popularity), which is low.  
  
  

Model performance metrics
```{r echo=T}
d1 <- data.frame(
  RMSE = caret::RMSE(reg.pcr.pred, data_test_regression$Popularity),
  Rsquare = caret::R2(reg.pcr.pred, data_test_regression$Popularity))
```

```{r echo=F}
kable(round(d1,2),caption = "Quality measures of the prediction using PCA",
    col.names = c("RMSE","R - Squared")) %>% 
kable_styling(latex_options = "hold_position",bootstrap_options = c("striped", "hover"),)

```
  
  
Performance metrics contained in Table 7 indicates that used independent variables are not explaining much in the variation of the dependent variable.
  
### Conclusion

From the performed analysis, we concluded that over 50% of the variation in the data, and as that there is no significant drop in explained variance with each additional variable.
Performing a Principal Component Regression on this problem revealed that the Popularity of songs is not strongly dependent on used independent variables. Further research on this topic could benefit from including more samples and from adding data from different periods to assess the importance of varying loading to principal components in different periods.  


[^1]: <https://www.forbes.com/sites/hughmcintyre/2019/04/02/the-global-music-industry-hits-19-billion-in-sales-in-2018-jumping-by-almost-10/#52c7da3e18a9>

[^2]: Dataset available at: <https://www.kaggle.com/leonardopena/top50spotify2019/data>
